---
title: "permafrost-black-carbon"
format: html
editor: visual
---







## Data

https://map.purpleair.com/1/mAQI/a10080/p604800/cC0?inc=698715&select=26423#13.58/68.62791/-149.60342

- black carbon

https://www.uaf.edu/toolik/edc/monitoring/abiotic/met-data-query.php

- soil temperature and weather

citation: 
Environmental Data Center Team. Year of data retrieval. Meteorological monitoring program at Toolik, Alaska. Toolik Field Station, Institute of Arctic Biology, University of Alaska Fairbanks, Fairbanks, AK 99775. https://www.uaf.edu/toolik/edc/monitoring/abiotic/met-data-query.php


## Load Data

Load packages and libraries needed to complete this analyis.

```{r, warning=FALSE}
library(here)
library(tidyverse)
library(lubridate)
library(readr)
library(gt)
library(tufte)
library(feasts)
library(tsibble)
library(sjPlot)
```

```{r, warning=FALSE}
# load data that contains pm2.5 which will be used as a representation of black carbon
black_carbon <- read_csv(here("data", "us-epa-pm25-aqi.csv"))

# load data that contains air temperature in degrees Celcius for Lake Toolik, Alaska
air_temp <- read_csv(here("data", "24-hour_data.csv"))

# load data that has soil temperature in degrees Celcius for first 150 cm
soil <- read_csv(here("data", "3-hour_data.csv"))
```

```{r}
# examin dataframes
head(air_temp)
head(soil)
head(black_carbon)
```

## Clean Data for Use

```{r}
# black carbon date information is stored as POSIXct so switch to date type
black_carbon$DateTime <- as.Date(black_carbon$DateTime) 


black_carbon_2 <- black_carbon %>% 
  rename(date = DateTime) %>%  #rename date column to match other df's
  mutate(b_c = (`Toolik Field Station A`+`Toolik Field Station B`)/2) %>% # average the two field station measurements
  mutate(log_b_c = (log(b_c)+1)) %>%  #b_c is not normally distributed so take log to correct for this
  subset(select = -c(Average,`Toolik Field Station A`, `Toolik Field Station B`))  # remove unecessary columns
  
black_carbon_2$log_b_c[black_carbon_2$log_b_c == -Inf] <- 0 # change -Inf values created by log function to 0 to allow regressions to run

# view edited data
head(black_carbon_2) 
```

```{r}
# create updated soil temperature df
soil_2 <- soil %>% 
  filter(hour == "300") %>% # choose all values recorded at hour 300
  subset(select = -hour) %>% # remove hour column
  rename(soil_temp = soil1_150cm) # rename column for readability

# view updated soil dataframe
head(soil_2)
```

```{r}
# create updated air temperature df
air_temp_updated <- air_temp %>% 
  select(air_temp_max_3m, date) %>% # select columns to keep
  rename(air_temp = air_temp_max_3m) # rename air_temp for readability
```

## Combine data

```{r}
# create new df containing black carbon and soil temp data
combined_bc_st <- left_join(black_carbon_2, soil_2, by = 'date')

# create new df with black carbon, soil temp, and air temp data
full <- left_join(combined_bc_st, air_temp_updated, by = 'date') %>% 
  drop_na() # remove na values 

# view edited df
head(full)
```

## Examine Data

```{r}
ggplot(data = full, aes(x = air_temp, y = soil_temp)) +
  geom_point() +
  theme_minimal() +
  geom_smooth(method = lm) +
  labs(x = "Air Temperature Degrees Celcius", y = "Soil Temperature Degrees Celcius")
```

```{r}
# plot black carbon and soil temp to see if linear regression is appropriate
ggplot(data = full, aes(x = b_c, y = soil_temp)) +
  geom_point() +
  theme_minimal() +
  geom_smooth(method = lm) +
  labs(x = "PM2.5 (ppm)", y = "Soil Temperature Degrees Celcius")

# clustering of data, use a histogram to examine the distribution
hist(full$b_c)
```

```{r}
# re-axamine black carbon with a log transformation applied
ggplot(data = full, aes(x = log_b_c, y = soil_temp)) +
  geom_point() +
  theme_minimal() +
  geom_smooth(method = lm) +
  labs(x = "PM2.5 (ppm)", y = "Soil Temperature Degrees Celcius")

# re-axamine black carbon with a log transformation applied
hist(full$log_b_c)
```

## Linear Regression

```{r}
# run linear regression on soil temp using air temp as predictor
permafrost_mod <- summary(lm(soil_temp ~ air_temp, data = full))
print(permafrost_mod)
```

```{r}
# run linear regression on soil temp using air temp and black carbon as predictors
permafrost_mod_2 <- summary(lm(soil_temp ~ air_temp + log_b_c, data = full))
print(permafrost_mod_2)
```

We use an interaction model to achieve this:

$$ soiltemp_i=\beta_{0}+\beta_{1} \cdot airtemp_i + \beta_{2} \cdot log(black carbon)_i + \beta_{3} \cdot airtemp_i \cdot log(blackcarbon)_i + \varepsilon_i$$

```{r}
# run linear regression on soil temp using air temp and black carbon as well as an interaction between black carbon and air temperature
permafrost_mod_3 <- lm(soil_temp ~ air_temp + log_b_c + air_temp:log_b_c, data = full)
permafrost_mod_3_summ <- summary(permafrost_mod_3)
print(permafrost_mod_3_summ)
```

```{r}
permafrost_table <- permafrost_mod_3 %>%
  tab_model()

permafrost_table
 # kbl(caption = "Recreating booktabs style table") %>%
 # kable_classic(full_width = F, html_font = "Cambria")
```

Hypothesis Testing 1: 

$$H_{0}:$$ There is no effect of air temperature on soil temperature. 
$$H_{A}:$$ There is an effect of air temperature on soil temperature. 

  - significant at 5%
  
Hypothesis Testing 2: 
$$H_{0}:$$ There is no effect of black carbon on soil temperature. 
$$H_{A}:$$ There is an effect of black carbon on soil temperature. 

  - significant at 10% level

Hypothesis Testing 3:

$$H_{0}:$$ there is no effects of air temperature that differs by blackcarbon levels on soil temperature
$$H_{A}:$$ There is an effect of air temperature and black carbon on soil temperature.
  
   - not significant
  
Interpretations : 

$$\beta_{0}$$ = When air temperature is 0 degrees Celcius and the black carbon level is 0 the predicted soil temperature is -1.74 degrees Celsius.

$$\beta_{1}$$ = There is a 0.06 predicted increase in soil temperature in degrees Celsius for every 1 degress celsius increase in air temperature holding black carbon fixed

$$\beta_{2}$$ = There is a 1 percent decrease predicted in soil temperature in degrees Celsius for every 0.0034 ppm increase in black carbon holding air temperature fixed.


0.06>0.0034 so black carbon is less significant than temperature

```{r}

```

## Additional Exploration

Does black carbon have a seasonal trend?

```{r}
permafrost <- as_tsibble(full) %>% 
  fill_gaps()

permafrost_2020 <- permafrost %>% 
  filter(between(date, as.Date('2019-12-29'), as.Date('2021-01-03')))

permafrost_2 <- permafrost_2020 %>% 
  mutate(date = date(date)) %>%
  ggplot(., aes(x = date, y = log_b_c)) +
  geom_line() +
  labs(x = 'Date', y = 'bc', title = 'black carbon') +
  theme_linedraw()

permafrost_2
```

```{r}
permafrost_3 <- permafrost_2020 %>% 
  model(STL(log_b_c, type = "additive")) %>% 
  components() %>% 
  autoplot()

permafrost_3
```

we have a large jump in our data occuring in septemper of 2020...there were major wildfires in the PNW in September 2020 that could be causing our models to overpredict

sources : https://journals.ametsoc.org/view/journals/wefo/36/5/WAF-D-21-0028.1.xml

let's try removing this event and seeing if our models are improved

## Remove

```{r}
permafrost_full <- permafrost %>% 
  filter(between(date, as.Date('2019-12-29'), as.Date('2023-11-19')))

permafrost_3 <- permafrost_full %>% 
  mutate(date = date(date)) %>%
  ggplot(., aes(x = date, y = log_b_c)) +
  geom_line() +
  labs(x = 'Date', y = 'bc', title = 'black carbon') +
  theme_linedraw()

permafrost_3
```
